In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-23 paper/main.tex
Numerical in main text

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Imposing a constant CSF production of
$400\,$ml/day across the surface of the lateral ventricles, while
allowing for efflux across the upper convexity, yields a total CSF
pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). To examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Methods). For the cardiac contribution, we compute
CSF pressure and flow fields at peak systolic blood inflow,
corresponding to a reduction in CSF space volume at a total rate of
$6\,$ml/s~\cite{baledent2014imaging, causemann2022human} in the SAS
and $0.31\,$ml/s across the lateral ventricle
surface\cite{vinje2019respiratory}. This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (see \Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs}D). The corresponding contribution to the periarterial
flow velocity is defined for each segment $\Lambda_i$ as $\langle Q'_i
\rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the cross-section
area of the PVS segment and $\langle Q_i' \rangle$ is the estimated
mean flow rate (see Supplementary Methods~\ref{sec:app:peristalsis}
for more details). Two wall motion patterns are considered, yielding
$\hat{u}_a^c$ and $\hat{u}_a^v$ respectively: \emph{cardiac}
pulsations with $f = 1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$
(\Cref{fig:pvs}E), and very low-frequency \emph{vasomotion} with $f =
0.1$Hz, $\lambda = 0.02$m, and $\varepsilon = 10\%$
(\Cref{fig:pvs}F). No flow induced by peristalsis is considered for
the perivenous network. In the baseline model, we combine the
contributions from CSF production and cardiac peristalsis to set
$\hat{u}$ in~\eqref{eq:multi_transport} as $\hat{u}_a = \hat{u}_a^p +
\hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v = \hat{u}_v^p$ in
$\Lambda_v$.

@@

\subsection*{Numerical approximation, implementation, verification}

A complete description of the numerical methods used is provided in
\Cref{sec:details_numerical_method}, while a condensed summary is
given here. To solve~\eqref{eq:multi_transport} numerically, we use
the interior penalty discontinuous Galerkin (DG) method with weighted
averages and upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low regularity properties. Thus, one can only expect lower-order error convergence near
$\Lambda_a$ and $\Lambda_v$, see \cite{masri2024discontinuous} for the numerical analysis of DG schemes and \cite{laurino2019derivation} for continuous Galerkin schemes.  However, away from these 1D networks, almost
optimal approximation is expected~\cite{masri2023discontinuous,koppl2016local}. To numerically solve the Stokes equations to
compute the CSF flow velocity and pressure in the ventricular system
and SAS, we use finite element spaces that preserve the
incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when later
solving~\eqref{eq:multi_transport} %\cite{cesmelioglu2022compatible}.
 The
spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension
FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}. The correctness
and accuracy of the numerical solutions were verified by a series of
numerical verification experiments \fixme{(see \Cref{sec:app:verification})}.

@@

As the primary means of model validation, we compare the in-silico
predictions of tracer enrichment and clearance against glymphatic MRI
studies (Discussion). In addition, we compare auxiliary model
quantities (CSF flow rates and pressure differences, dispersion factor
estimates, and the shapes and sizes of PVSs) with the current
literature (Supplementary discussion: Extended model validation).

>>>>

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Imposing a constant CSF production of
$400\,$ml/day across the surface of the lateral ventricles, while
allowing for efflux across the upper convexity, yields a total CSF
pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). To examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Methods). For the cardiac contribution, we compute
CSF pressure and flow fields at peak systolic blood inflow,
corresponding to a reduction in CSF space volume at a total rate of
$6\,$ml/s~\cite{baledent2014imaging, causemann2022human} in the SAS
and $0.31\,$ml/s across the lateral ventricle
surface\cite{vinje2019respiratory}. This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs}D). The corresponding contribution to the periarterial
flow velocity is defined for each segment $\Lambda_i$ as $\langle Q'_i
\rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the cross-section
area of the PVS segment and $\langle Q_i' \rangle$ is the estimated
mean flow rate (see~\ref{sec:app:peristalsis} for more details). Two
wall motion patterns are considered, yielding $\hat{u}_a^c$ and
$\hat{u}_a^v$ respectively: \emph{cardiac} pulsations with $f =
1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$ (\Cref{fig:pvs}E), and
very low-frequency \emph{vasomotion} with $f = 0.1$Hz, $\lambda =
0.02$m, and $\varepsilon = 10\%$ (\Cref{fig:pvs}F). No flow induced by
peristalsis is considered for the perivenous network. In the baseline
model, we combine the contributions from CSF production and cardiac
peristalsis to set $\hat{u}$ in~\eqref{eq:multi_transport} as
$\hat{u}_a = \hat{u}_a^p + \hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v
= \hat{u}_v^p$ in $\Lambda_v$.

@@

\subsection*{Numerical approximation, implementation, and verification}

We give a concise summary of the numerical approach here (see
Supplementary methods for a comprehensive description). To
solve~\eqref{eq:multi_transport} numerically, we use the interior
penalty discontinuous Galerkin (DG) method with weighted averages and
upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low
regularity properties. Thus, one can only expect lower-order error
convergence near $\Lambda_a$ and $\Lambda_v$, see the references
\cite{masri2024discontinuous} and \cite{laurino2019derivation} for the
numerical analysis of DG and continuous Galerkin schemes,
respectively. However, away from these 1D networks, almost optimal
approximation is
expected~\cite{masri2023discontinuous,koppl2016local}. To numerically
solve the Stokes equations to compute the CSF flow velocity and
pressure in the ventricular system and SAS, we use finite element
spaces that preserve the incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when subsequently
solving~\eqref{eq:multi_transport}\cite{cesmelioglu2022compatible}.
The spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension
FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}. The correctness
and accuracy of the numerical solutions were verified by a series of
numerical verification experiments (\Cref{sec:app:verification}).

@@

As the primary means of model validation, we compare the in-silico
predictions of tracer enrichment and clearance against glymphatic MRI
studies (Discussion). In addition, we compare auxiliary model
quantities (CSF flow rates and pressure differences, dispersion factor
estimates, and the shapes and sizes of PVSs) with the current
literature (\Cref{sec:app:model_validation}).

### 2025-01-23 paper/supplementary_main.tex
Numerical in main text

\subsubsection{Modelling CSF flow via the incompressible Stokes equations}
\label{sec:csf_flow}
The time-independent, incompressible Stokes equations model the flow
of an incompressible Newtonian fluid at low Reynolds numbers, and read
as follows: over the domain $\Omega_{\rm CSF} \subset \R^3$, find the
velocity vector field $\bm u : \Omega_{\rm CSF} \rightarrow \R^3$ and
the pressure field $p : \Omega_{\rm CSF} \rightarrow \R$ such that
\begin{subequations}
  \begin{alignat}{2}
    - \mu \Delta \bm u  + \nabla p &=  0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:momentum_equation} \\ 
    \nabla \cdot \bm u &= 0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:divergence_equation} .
  \end{alignat}
  \label{eq:stokes}%
\end{subequations}%
In addition, we impose the following boundary conditions to model flow
induced by CSF production in the choroid plexus with $\Gamma_{\rm AM-U}$ as the main efflux site:
\begin{subequations}
  \begin{alignat}{2}
    \mu \nabla \bm u \cdot \bm{n} - p \bm n &= -R_0 ( \bm u \cdot \bm n ) \bm n
    && \text{ on } \Gamma_{\mathrm{AM-U}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\mathrm{AM-L}}\cup\Gamma_{\rm{pia}}\cup\Gamma_{\rm{SSAS}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{pia}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{SSAS}}, \\
    \bm u \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \text{ on } \Gamma_{\rm{LV}} ,  
  \end{alignat}
  \label{eq:stokes_bcs}%
\end{subequations}%
where $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the (dynamic) CSF viscosity, $R_0 > 0$ is a resistance parameter for CSF efflux, and $u_{\rm in}$ is a given fluid influx, here across the lateral ventricle wall $\Gamma_{\rm LV}$. We consider alternative variations of these boundary conditions in connection with estimating dispersion effects induced by CSF pulsatility, see~\Cref{sec:app:dispersion}.

@@

\subsection{Numerical solution of the transport and flow equations}
\label{sec:details_numerical_method}

@@

\subsection{Numerical verification}

@@

\section{Supplementary discussion}
\label{sec:model_validation}

\subsection{Extended model validation}
\label{sec:model_validation}

>>>>

\subsubsection{Modelling CSF flow via the incompressible Stokes equations}
\label{sec:app:csf_flow}
The time-independent, incompressible Stokes equations model the flow
of an incompressible Newtonian fluid at low Reynolds numbers, and read
as follows: over the domain $\Omega_{\rm CSF} \subset \R^3$, find the
velocity vector field $\bm u : \Omega_{\rm CSF} \rightarrow \R^3$ and
the pressure field $p : \Omega_{\rm CSF} \rightarrow \R$ such that
\begin{subequations}
  \begin{alignat}{2}
    - \mu \Delta \bm u  + \nabla p &=  0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:momentum_equation} \\ 
    \nabla \cdot \bm u &= 0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:divergence_equation} .
  \end{alignat}
  \label{eq:stokes}%
\end{subequations}%
In addition, we impose the following boundary conditions to model flow
induced by CSF production in the choroid plexus with $\Gamma_{\rm AM-U}$ as the main efflux site:
\begin{subequations}
  \begin{alignat}{2}
    \mu \nabla \bm u \cdot \bm{n} - p \bm n &= -R_0 ( \bm u \cdot \bm n ) \bm n
    && \text{ on } \Gamma_{\mathrm{AM-U}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\mathrm{AM-L}}\cup\Gamma_{\rm{pia}}\cup\Gamma_{\rm{SSAS}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{pia}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{SSAS}}, \\
    \bm u \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \text{ on } \Gamma_{\rm{LV}} ,  
  \end{alignat}
  \label{eq:stokes_bcs}%
\end{subequations}%
where $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the (dynamic) CSF viscosity, $R_0 > 0$ is a resistance parameter for CSF efflux, and $u_{\rm in}$ is a given fluid influx, here across the lateral ventricle wall $\Gamma_{\rm LV}$. We consider alternative variations of these boundary conditions in connection with estimating dispersion effects induced by CSF pulsatility, see~\Cref{sec:app:dispersion}.

@@

\subsection{Numerical solution of the transport and flow equations}
\label{sec:app:details_numerical_method}

@@

\subsection{Numerical verification}
\label{sec:app:verification}

@@

\section{Supplementary discussion}

\subsection{Extended model validation}
\label{sec:app:model_validation}
